smooth-sea

### 2018-05-19 index.rmd
some progress on index

#source("./smoothSpatiotemporal.R", echo = FALSE)
```

@@

>>>>

#packages
library(tidyverse)
library(forcats)
library(haven)
library(lubridate)
library(haven)
library(zoo)
library(ggthemes)
library(RColorBrewer)
library(sp)
library(spdplyr)
library(geosphere)
library(rgdal)
library(rgeos)
library(sqldf)
library(spdep)
library(viridis)
library(latex2exp)
library(plotly)

#load ACS data extract (codebook in folder)
census <- read_csv(file = "./input/acsExtract.csv")

#read in tract shapefiles for seattle
sea_shp <- readOGR(dsn = "R:/Project/seattle_rental_market/data/geo/sea_tract_2010/sea_tract_2010.shp",
                  layer = "sea_tract_2010",
                  GDAL1_integer64_policy = TRUE,
                  stringsAsFactors = F)

#determine config (dev or github)
if(file.exists("../data/cl/craigslistDB.sqlite")){
  #connect to database if file.exists
  DB <- dbConnect(SQLite(), dbname="../data/cl/craigslistDB.sqlite")
  cl <- tbl(DB, "clean") #create dbi for clean listing table

  #compute tract aggregates for CL listing count
  tract <- cl %>%
    collect %>% #bring db query into memory
    filter(!is.na(GISJOIN), !is.na(cleanBeds), !is.na(cleanRent), !is.na(cleanSqft), 
           GISJOIN %in% sea_shp@data$GISJOIN) %>% #only listings with valid Bed/Rent, seattle tracts
    distinct(cleanBeds, cleanRent, cleanSqft, matchAddress, matchAddress2, .keep_all = T) %>% #dedupe to unique address-bed-rent combos
    dplyr::select(listingDate, GEOID10, GISJOIN, seattle, matchAddress, matchType, cleanBeds, cleanRent, cleanSqft) %>% #SELECT these columns
    mutate(listingDate = as.Date(listingDate),
           listingQtr = as.yearqtr(listingDate)) %>%
    filter(cleanBeds %in% c(0, 1, 2, 3)) %>%
    group_by(GEOID10, GISJOIN, listingQtr) %>% #group listings by tract, qtr within tract
    summarize(nListings = n(),
              n1B = sum(cleanBeds == 1),
              nGT2B = sum(cleanBeds > 2),
              medRent = median(cleanRent),
              med0B = median(cleanRent[cleanBeds==0]),
              med1B = median(cleanRent[cleanBeds==1]),
              med2B = median(cleanRent[cleanBeds==2]),
              med3B = median(cleanRent[cleanBeds==3]),
              lagRent = lag(cleanRent)) %>% #create tract aggregates
    mutate(pGT2B = nGT2B/nListings) %>%
    ungroup %>%
    arrange(GISJOIN, listingQtr)
  dbDisconnect(DB)
  write_csv(tract, "./input/tractCl.csv")

} else{
  #if github config, read in extract from cl db
  tract <- read_csv("./input/tractCl.csv")
}

#### A. construct panel (some missingness) ------------------------------------

census <- census[match(sea_shp@data$GISJOIN, census$GISJOIN),]

census <- census %>%
  mutate(nHU = AF7PE001,
         pOwnoccHU = AF7PE002/AF7PE001,
         medHUVal = AF9LE001)

### Training periods

#studios
Q2_2017 <- tract %>%
  filter(listingQtr == "2017 Q2") %>%
  right_join(census) %>%
  ungroup %>%
  mutate(listingQtr = "2017 Q2",
         Qtr = "Q2",
         actualRent = med1B)

Q3_2017 <- tract %>%
  filter(listingQtr == "2017 Q3") %>%
  right_join(census) %>%
  ungroup %>%
  mutate(listingQtr = "2017 Q3",
         Qtr = "Q3",
         actualRent = med1B)

Q4_2017 <- tract %>%
  filter(listingQtr == "2017 Q4") %>%
  right_join(census) %>%
  ungroup %>%
  mutate(listingQtr = "2017 Q4",
         Qtr = "Q4",
         actualRent = med1B)

Q1_2018 <- tract %>%
  filter(listingQtr == "2018 Q1") %>%
  right_join(census) %>%
  ungroup %>%
  mutate(listingQtr = "2018 Q1",
         Qtr = "Q1",
         actualRent = med1B)

### Forecast/test period

Q2_2018 <- tract %>%
  filter(listingQtr == "2018 Q2") %>%
  right_join(census) %>%
  ungroup %>%
  mutate(listingQtr = "2018 Q2",
         Qtr = "Q2",
         actualRent = med1B,
         medRent = NA)
Q2_2018 <- Q2_2018[match(sea_shp@data$GISJOIN, Q2_2018$GISJOIN),]

sea_df <- bind_rows(Q2_2017, Q3_2017, Q4_2017, Q1_2018, Q2_2018)
sea_df$listingQtr <- as.yearqtr(sea_df$listingQtr)

sea_df <- sea_df %>% 
  dplyr::select(-GEOID10) %>%
  dplyr::select(GISJOIN, listingQtr, Qtr, medRent, med1B, lagRent, actualRent, nHU, 
                pOwnoccHU, medHUVal, pGT2B)

#### ACF for panel ------------------------------------------------------------

acf_df <- data.frame(sea_df = unique(sea_df$GISJOIN), acf = rep(NA, length(unique(sea_df$GISJOIN))))
for(i in unique(sea_df$GISJOIN)){
  obs <- sea_df[sea_df$GISJOIN == i,]
  ts <- ts(obs$medRent)
  if(sum(is.na(ts)) > 0){
    next
  } else{
    #print(ts)
    acf <- acf(ts)
    #print(acf)
    acf_df$acf[acf_df$sea_df == tract] = list(acf)
  }
}

#ACFs are weak with only 5 periods to assess with. Highest L1 value is around .5 but many are
#weakened by a noisey quarter

#### F. INLA Models of CL Rent ------------------------------------------------

library(INLA)

#geo information for INLA
sea_shp <- readOGR(dsn = "R:/Project/seattle_rental_market/data/geo/sea_tract_2010/sea_tract_2010.shp",
                   layer = "sea_tract_2010",
                   GDAL1_integer64_policy = TRUE,
                   stringsAsFactors = F)

cens <- gCentroid(sea_shp, byid = T)
cens <- data.frame(long = cens@coords[,1],
                   lat = cens@coords[,2])
cens <- cens %>% mutate(row = row_number())
sea_ff <- fortify(sea_shp)

ggplot(sea_ff, aes(x = long, y = lat, group = group)) + 
  geom_polygon(fill = "grey90", color = "white") +
  geom_text(data = cens, aes(x  = long, y = lat, label = row, group = row))

sea_adj <- poly2nb(sea_shp)
nb2INLA("R:/Project/seattle_rental_market/report/spatial_epi/seatract.graph", sea_adj)
sea_df$idtract <- rep(1:134, 5)
sea_df$idtract1 <- rep(1:134, 5)
sea_df$idqtr <- as.numeric(sea_df$listingQtr)
sea_df$idqtr1 <- sea_df$idqtr
sea_df$idtractqtr <- paste(sea_df$idtract, sea_df$idqtr)

#Seattle median rent models -----------------------------------------------

#fixed effect for qtr only
form.int <- med1B ~ 1 + Qtr

m.int <- inla(form.int, 
             family = "lognormal", 
             data = sea_df,
             control.predictor = list(compute = TRUE),
             control.compute = list(dic = TRUE, waic = TRUE))

summary(m.int)
sea_df$int_Med <- m.int$summary.linear.predictor[, "0.5quant"]
sea_df$int_SD <- m.int$summary.linear.predictor[, "sd"]
sea_df$int_postWidth <- m.int$summary.linear.predictor[, "0.975quant"] - m.int$summary.linear.predictor[,"0.025quant"]
sea_df$int_Eff <- m.int$summary.random$idtract[sea_df$idtract, "0.5quant"]

#fixed qtr + iid tract random effect
form.ns <- med1B ~ 1 + Qtr + 
   f(idtract, model = "iid")

m.ns <- inla(form.ns, 
              family = "lognormal", 
              data = sea_df,
              control.predictor = list(compute = TRUE),
              control.compute = list(dic = TRUE, waic = TRUE))

summary(m.ns)
sea_df$ns_Med <- m.ns$summary.linear.predictor[, "0.5quant"]
sea_df$ns_SD <- m.ns$summary.linear.predictor[, "sd"]
sea_df$ns_postWidth <- m.ns$summary.linear.predictor[, "0.975quant"] - m.ns$summary.linear.predictor[,"0.025quant"]
sea_df$ns_Eff <- m.ns$summary.random$idtract[sea_df$idtract, "0.5quant"]

#IID tract random effect + AR(1) process
form.nsar1 <- med1B ~ 1 +
  f(idtract, model = "iid") +
  f(idqtr, model = "ar1") + f(idqtr1, model = "iid")

m.nsar1 <- inla(form.nsar1, 
               family = "lognormal", 
               data = sea_df,
               control.predictor = list(compute = TRUE),
               control.compute = list(dic = TRUE, waic = TRUE))

summary(m.nsar1) #NB: does not improve fit to use random effect time specification, using fixed effects
sea_df$nsar1_Med <- m.nsar1$summary.linear.predictor[, "0.5quant"]
sea_df$nsar1_SD <- m.nsar1$summary.linear.predictor[, "sd"]
sea_df$nsar1_postWidth <- m.nsar1$summary.linear.predictor[, "0.975quant"] - m.nsar1$summary.linear.predictor[,"0.025quant"]
sea_df$nsar1_Eff <- m.nsar1$summary.random$idtract[sea_df$idtract, "0.5quant"]

#spatial random effect model (BYM 1991) with seasonal dummies
form.bym <- med1B ~ 1 + Qtr +
  f(idtract, model = "bym", #ICAR spatial RE for tract neighbors + IID RE for tract
    scale.model = T,
    graph = "R:/Project/seattle_rental_market/report/spatial_epi/seatract.graph")

m.bym <- inla(form.bym, 
             family = "lognormal", 
             data = sea_df,
             control.predictor = list(compute = TRUE),
             control.compute = list(dic = TRUE, waic = TRUE))

summary(m.bym)
sea_df$bym_Med <- m.bym$summary.linear.predictor[, "0.5quant"]
sea_df$bym_SD <- m.bym$summary.linear.predictor[, "sd"]
sea_df$bym_postWidth <- m.bym$summary.linear.predictor[, "0.975quant"] - m.bym$summary.linear.predictor[,"0.025quant"]
sea_df$bym_Eff <- m.bym$summary.random$idtract[sea_df$idtract, "0.5quant"]

#spatio-temporal smoothing model with linear space-time interaction (Bernadelli 1995)
form.spt <- med1B ~ 1 + Qtr +
  f(idtract, model = "bym", #ICAR spatial RE for tract neighbors + IID RE for tract
    scale.model = T,
    graph = "R:/Project/seattle_rental_market/report/spatial_epi/seatract.graph") +
  f(idtractqtr, model="iid") + #exchangeable RE for period
  f(idqtr, model = "iid")

m.spt <- inla(form.spt, 
              family = "lognormal", 
              data = sea_df,
              control.predictor = list(compute = TRUE),
              control.compute = list(dic = TRUE, waic = TRUE))

summary(m.spt)
sea_df$spt_Med <- m.spt$summary.linear.predictor[, "0.5quant"]
sea_df$spt_SD <- m.spt$summary.linear.predictor[, "sd"]
sea_df$spt_postWidth <- m.spt$summary.linear.predictor[, "0.975quant"] - m.spt$summary.linear.predictor[,"0.025quant"]
sea_df$spt_Eff <- m.spt$summary.random$idtract[sea_df$idtract, "0.5quant"]

#overfitting? too much bumpiness qtr-to-qtr

#### Maps of Model Output ------------------------------------------------------------------

sea_df <- sea_df %>%
  mutate_at(.vars = vars(matches("_Med")),
            .funs = exp) %>%
  mutate_at(.vars = vars(matches("_SD")),
            .funs = exp) %>%
  group_by(GISJOIN) %>% 
  arrange(GISJOIN, listingQtr) %>%
  mutate(spt_traj = spt_Med - lag(spt_Med, 4)) %>%
  ungroup %>%
  group_by(GISJOIN) %>%
  mutate(spt_traj = sum(spt_traj, na.rm=T))

sea_df

sea_shp@data$id <- rownames(sea_shp@data)
sea_shp@data <- left_join(sea_shp@data, sea_df)
sea_f <- fortify(sea_shp)
sea_f <- inner_join(sea_f, sea_shp@data,"id")

rmse <- function(error){sqrt(mean(error^2))}
mae <- function(error){mean(abs(error))}

#table of fit statistics
fit_stats <- sea_df %>%
  mutate(train_test = ifelse(listingQtr == "2018 Q2", "Test", "Training")) %>%
  mutate(int_err = actualRent - int_Med,
         ns_err = actualRent - ns_Med,
         nsar1_err = actualRent - nsar1_Med,
         bym_err = actualRent - bym_Med,
         spt_err = actualRent - spt_Med) %>% 
  group_by(train_test) %>%
  summarize(int_rmse = rmse(int_err[!is.na(int_err)]),
            int_mae = mae(int_err[!is.na(int_err)]),
            ns_rmse = rmse(ns_err[!is.na(ns_err)]),
            ns_mae = mae(ns_err[!is.na(ns_err)]),
            nsar1_rmse = rmse(nsar1_err[!is.na(nsar1_err)]),
            nsar1_mae = mae(nsar1_err[!is.na(nsar1_err)]),
            bym_rmse = rmse(bym_err[!is.na(bym_err)]),
            bym_mae = mae(bym_err[!is.na(bym_err)]),
            spt_rmse = rmse(spt_err[!is.na(spt_err)]),
            spt_mae = mae(spt_err[!is.na(spt_err)])) %>%
  dplyr::select(train_test, ends_with("_rmse"), ends_with("_mae"))

```

@@

### Observed vs. Smoothed Rent Estimates

```{r index.maps, echo=F}
smoothed <- ggplot(sea_f %>% filter(listingQtr == "2018 Q2"), 
                   aes(x = long, y = lat, group = group, fill = bym_Med, label = GISJOIN)) +
  #facet_grid(~ listingQtr) +
  geom_polygon(color = "grey85", lwd = .05) +
  scale_fill_viridis_c(limits = c(750, 2500), labels = scales::dollar) +
  coord_quickmap() +
  theme(axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank()) +
  labs(title = "Spatially Smoothed 1B Median Rent",
       fill = "Smoothed")
ggplotly(smoothed, tooltip = c("label", "fill"))
```
